### 2020-03-26 HICSSSubmission/bp_results_1.tex
reducing text

To gain insights into how knowledge workers experience stress over an
extended period of work, we examined the end of day survey responses
collected from each participant to see if any identifiable trends
emerged. We used data from 13 of the 14 participants - we excluded one participant
from this analysis as they experienced atypical stress
levels in the latter half of the study due to factors outside of our
control.

@@

Common amongst all participants was a trend to select one stress rating
far more frequently than any other. We will refer to this value as the
participant's baseline stress level. All but one participant reported
their perceived stress level for the day as their baseline stress
level more than 50\% of the time. In total, the baseline values made
up 65\% of the reported values collected from
participants. Interestingly, while participants sometimes saw periods
of sustained increases in stress, lasting as many as 6 consecutive
workdays in the most extreme case, participants would always return to
their baseline stress level at some point.

\begin{figure}
  \centering
      \includegraphics[width=0.4\textwidth]{dailystress_s1.png}
      \includegraphics[width=0.4\textwidth]{dailystress_s3.png}
  \caption{The day-to-day stress levels reported by two participants (S1 and S3) are shown. The y-axis represents values on the 5-point Likert scale we asked participants to respond with ranging from 1/Not at all stressed to 5/Extremely stressed. The x-axis represents the day of the study on which the value was recorded (from 0-45). The x-axes are slightly different for the 2 participants as they did not begin and end the study on exactly the same days.}
   \label{fig:dailyStress}
   \vspace*{-2mm}
\end{figure}

@@

Accounting for the variance between participants perceived stress
baselines, we consider a stressful day to be one that represents a
deviation of 1 or more stress levels above the participant's
baseline. Of the 93 stressful days we observed in total, we found that
39 (41\%) of these days occurred in groupings of two or more
consecutive stressful workdays. The most common size of these groups
was two workdays, while the largest group we observed was six
workdays.  The day after a stressful day is much more likely to be a
stressful day as compared to any other day with a 0.55 average increase
over baseline, compared to 0.02 average increase over baseline.

\subsection{Extreme Changes in Stress Levels are Rare}

After accounting for each participant's perceived stress baseline, we
examined the frequency of deviations from the baseline. We found that
participants were far more likely to report a stress level that was
within 1 point of their baseline, than to report a stress level 2 or
more points away. These extreme deviations represented only 15\% of
all reported values which differed from the participants baseline. As
well, the majority (78\%) of these deviations came from just two
participants. This suggests that some people may be less resilient to
the stress of the workplace than others. For most participants,
extremely stressful days were few and far between.

@@

In an attempt to explain some of the stress that our participants were experiencing, we created a linear mixed model with the self-reported daily stress levels as dependent variable and the participants as random effects. We experimented with day of the week and proximity to beginning or end of month as possible explanatory variables. We also calculated daily features from the gathered computer interaction data where available---namely total time spent actively on the computer, percentage of time idle, and amount of time spent on non work-related web browsing per day. Ultimately, the analysis showed that none of the variables that we examine had a significant explanatory power with respect to our participants perceived stress levels. This indicates that aspects such as the time spent on a computer or the non work-related web browsing do not have a significant impact on the stress level. Furthermore, this points to the need for additional instrumentation and data collection, if we are to succesfully understand and predict stress in the workplace.

@@

To investigate whether stress, focus and awakeness can be predicted in
the moment based on biometric measures,
we investigated classifiers trained
for each individual and across all participants. We report on the effectiveness of
these classifiers and the features that are important in prediction of stress, focus
and awakeness.

@@

In this section we describe how we preprocessed the collected data for use in training and testing machine learning models.

@@

We linked the collected biometric data and survey responses for each participant. Linking the data is necessary to construct training and test datasets for use in creating and evaluation machine learning models.

@@

Reasons for a survey response to lack associated biometric data include:
\begin{itemize}
\item The participant not wearing the Everion in the hour before before beginning the survey
\item The Everion not recording data in the hour before the participant began the survey (e.g., due to low battery)
\item Biometric data not being uploaded successfully to the server
%\item Compatibility issues between the sensor and the OS tracking the participant's data
%\item Issues accessing the data uploaded by the participants
\end{itemize}

@@

\begin{figure}
  \centering
      \includegraphics[width=0.5\textwidth]{DuringTheDay.png}
  \caption{The figure shows the biometric data available per each participant. Green sections represent survey responses with biometric
  data available, red sections responses with no biometric data available}
   \label{surveyBio}
   \vspace*{-4mm}
\end{figure}

@@

We extracted features from the biometric data to provide as input to machine 
learning models. Previous 
studies~ \cite{vorburger05,zuger2015interruptibility} identify time windows as an 
important factor that impacts the prediction accuracy of a classifier. We 
considered many time windows from the literature on biometric 
analysis~\cite{zuger18}, ranging from 10 seconds to 3 hours. Specifically, we 
considered the following time windows: \textit{10sec, 20sec, 30sec, 45sec, 
1min, 2min, 3min, 5min, 7.5min, 10min, 20min, 30min, 45min, 1hour, 2hour, 
3hour}.

@@

Table~\ref{responseDistribution} illustrates the distribution of responses from each participant for each of the three survey questions (listed in Section~\ref{sec:Surveys}). The figure shows that there is a notable imbalance in the distribution of the self-reported responses provided by the participants. Most participants did not use all five points of the five-point Likert scale in their responses, and the distributions tend to skew toward one side or the other, depending on the question. Thus, we binarized the survey data into a two-point scale to give the machine learning models the best possible chance to make useful predictions. The two points in the binary scale represent negative or positive responses for each of the three human aspects of interests (e.g., not stressed or stressed).

@@

Even after binarizing the responses as described in the previous section, we found the distribution of responses was still quite imbalanced for many of our participants. This can be seen in the distribution columns in Table \ref{tab:accuracy}. To combat this, we applied random oversampling to our training sets, which artificially rebalances the dataset by creating randomly replicated data in the minority class. This has been a commonly used technique in previous studies on unbalanced datasets \cite{chawla2004,yap2014}.

\begin{table}
  \centering
      \includegraphics[width=0.5\textwidth]{distributiontable.pdf}
  \caption{The distribution of the responses of each participant to the three questions asked during the day are shown. Each bar in the histograms represent one of the 5 values on the 5-point Likert scale we asked participants to respond with, where the far left side of the histograms are 1/Not at all, and the far right sides are 5/Extremely}
   \label{responseDistribution}
   \vspace*{-2mm}
\end{table}

@@

\subsection{Selecting a Classifier Algorithm}

There are many different algorithms that can be used to build a classifier. To select an
algorithm, we compared multiple classifiers using the popular machine learning library scikit-learn~\cite{pedregosa11} and performed a grid search analysis to determine the optimal hyperparameters for each classifier. Our analysis showed that random forest outperforms all other classifiers, including Na\"ive Bayes, decision trees, support vector machine, and a multilayer perceptron neural network. The optimal values for random forest and the three output measures are listed in Table \ref{tab:hyperparams}. For the remainder of this paper, we refer to random forest classifiers trained with these hyperparameters.%\\[-0.1cm]
% for stress they were (minimum samples for split = 4, # of estimators = 100, max features = 0.5, k=300), for focus they were (minimum samples for split = 4, # of estimators = 50, max features = 0.5, k=200), and for awakeness they were (minimum samples for split = 4, # of estimators = 100, max features = 0.25, k=800)

@@

%\noindent\textit{Results of Individual Classifiers}\\
Since peoples' experience of stress, focus, and awakeness (as well as
their physiological manifestations) can vary substantially
(e.g.,~\cite{Hernandez11}), we first trained individual classifiers
for each participant (as opposed to a general one for all
participants). The results of our analysis are reported in
Table~\ref{tab:accuracy}. For our analysis, we report values of
accuracy, one of the most commonly used metric to compare performance,
as well as precision and recall of the classes of interest:
'stressed', 'not focused', and 'not awake'. Since the imbalance in the
data can lead to high accuracy values if a classifier always just
predicts the most likely/frequent class while ignoring the class of
higher importance and interest, precision and recall of the class of
interest are also important to
consider~\cite{yap2014,bhattacharyya_data_2011,Hernandez11}.
For some users (i.e., S1. as seen in Table \ref{responseDistribution}), the imbalance in their data was so extreme that even after adjusting by oversampling it was impossible to create a reasonable classifier. These scenarios are difficult to predict as any classifier will not have enough variance in its training data for the 'stressed' situation to adequately distinguish it from the non-stressed case.

@@

%\noindent\textit{Feature Selection and Importance}\\
There are a large variety of features that can be (and have been) calculated in previous research for each of the basic measurements listed in Table~\ref{signals}, such as the mean, standard deviation, maximum, and interquartile range. In addition, each of these metrics can be combined with the various time windows captured of a basic measurement, resulting in a large feature space. To reduce the feature space, we experimented with multiple feature selection methods, including selecting the top k highest correlated features by various metrics such as mutual information, Pearson's correlation coefficient, ANOVA's F-value, as well as wrapper methods such as recursive feature elimination, optimizing mean decrease accuracy by iteratively permuting features, and only selecting features that exceed a certain Gini importance threshold. We found that all methods produced similar results with respect to accuracy, precision, and recall for the individual models. Ultimately, we chose to use the top k features with the highest ANOVA F-value, as it is relatively simple and efficient to calculate. The values of k used were selected by grid search analysis, and are shown in Table \ref{tab:hyperparams}.

@@

%\noindent\textit{Individual vs.\ General Model}\\
Individual models are trained specifically for each individual and thus require a data collection period before they are capable of making accurate predictions. On the other hand, the idea of general models is to be able to train them on already collected data and then  to be able to apply them even to new and unseen individuals, thus overcoming the cold-start problem. Given the large individual differences in biometrics, training a general model to achieve an adequate accuracy for new individuals is not necessarily possible.

>>>>

\vspace{-4mm}
To gain insights into how knowledge workers experience stress over an
extended period of work, we examined the end of day survey responses
collected from each participant to see if any identifiable trends
emerged. We used data from 13 of the 14 participants - we excluded one participant
from this analysis as they experienced atypical stress
levels in the latter half of the study due to factors outside of our
control.

@@

\vspace{-4mm}
Common among all participants was a trend to select one stress rating
far more frequently than any other. We will refer to this value as the
participant's baseline stress level. All but one participant reported
their perceived stress level for the day as their baseline stress
level more than 50\% of the time. In total, the baseline values made
up 65\% of the reported values collected from
participants. Interestingly, while participants sometimes saw periods
of sustained increases in stress, lasting as many as 6 consecutive
workdays in the most extreme case, participants would always return to
their baseline stress level at some point.

\begin{figure}
  \centering
      \includegraphics[width=0.4\textwidth]{dailystress_s1.png}
      \includegraphics[width=0.4\textwidth]{dailystress_s3.png}
  \caption{The day-to-day stress levels reported by two participants. Y-axis represents values on the 5-point Likert scale. The x-axis represents the day of the study.}
   \label{fig:dailyStress}
   \vspace*{-2mm}
\end{figure}

@@

\vspace{-4mm}
Accounting for the variance between participants perceived stress
baselines, we consider a stressful day to be one that represents a
deviation of 1 or more stress levels above the participant's
baseline. Of the 93 stressful days we observed in total, we found that
39 (41\%) of these days occurred in groupings of two or more
consecutive stressful workdays. The most common size of these groups
was two workdays, while the largest group we observed was six
workdays.  The day after a stressful day is much more likely to be a
stressful day as compared to any other day with a 0.55 average increase
over baseline, compared to 0.02 average increase over baseline.

\subsection{Extreme Changes are Rare}

\vspace{-4mm}
After accounting for each participant's perceived stress baseline, we
examined the frequency of deviations from the baseline. We found that
participants were far more likely to report a stress level that was
within 1 point of their baseline, than to report a stress level 2 or
more points away. These extreme deviations represented only 15\% of
all reported values which differed from the participants baseline. As
well, the majority (78\%) of these deviations came from just two
participants. This suggests that some people may be less resilient to
the stress of the workplace. For most participants,
extremely stressful days were few and far between.

@@

\vspace{-4mm}
In an attempt to explain some of the stress that our participants were experiencing, we created a linear mixed model with the self-reported daily stress levels as dependent variable and the participants as random effects. We experimented with day of the week and proximity to beginning or end of month as possible explanatory variables. We also calculated daily features from the gathered computer interaction data where available---namely total time spent actively on the computer, percentage of time idle, and amount of time spent on non work-related web browsing per day. Ultimately, the analysis showed that none of the variables that we examine had a significant explanatory power with respect to our participants perceived stress levels. This indicates that aspects such as the time spent on a computer or the non work-related web browsing do not have a significant impact on the stress level. Furthermore, this points to the need for additional instrumentation and data collection, if we are to successfully understand and predict stress in the workplace.

@@

\vspace{-4mm}
To investigate whether stress, focus and awakeness can be predicted in
the moment based on biometric measures,
we investigated classifiers trained
for each individual and across all participants. We report on the effectiveness of
these classifiers and the features that are important in prediction of stress, focus
and awakeness.

@@

\vspace{-4mm}
In this section we describe how we preprocessed the collected data for use in training and testing machine learning models.

@@

\vspace{-8mm}
We linked the collected biometric data and survey responses for each participant. Linking the data is necessary to construct training and test datasets for use in creating and evaluation machine learning models.

@@

Reasons for a survey response to lack associated biometric data include:
\begin{itemize}[topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
\item The participant not wearing the Everion in the hour before before beginning the survey
\item The Everion not recording data in the hour before the participant began the survey (e.g., due to low battery)
\item Biometric data not being uploaded successfully to the server
%\item Compatibility issues between the sensor and the OS tracking the participant's data
%\item Issues accessing the data uploaded by the participants
\end{itemize}

@@

\begin{figure}
  \centering
      \includegraphics[width=0.5\textwidth]{DuringTheDay.png}
  \caption{Biometric data available per each participant. Green sections represent survey responses with biometric
  data available, red sections responses with no biometric data available}
   \label{surveyBio}
   \vspace*{-4mm}
\end{figure}

@@

\vspace{-8mm}
We extracted features from the biometric data to provide as input to machine 
learning models. Previous 
studies~ \cite{vorburger05,zuger2015interruptibility} identify time windows as an 
important factor that impacts the prediction accuracy of a classifier. We 
considered many time windows from the literature on biometric 
analysis~\cite{zuger18}, ranging from 10 seconds to 3 hours. Specifically, we 
considered the following time windows: \textit{10sec, 20sec, 30sec, 45sec, 
1min, 2min, 3min, 5min, 7.5min, 10min, 20min, 30min, 45min, 1hour, 2hour, 
3hour}.

@@

\vspace{-8mm}
Table~\ref{responseDistribution} illustrates the distribution of responses from each participant for each of the three survey questions (listed in Section~\ref{sec:Surveys}). The figure shows that there is a notable imbalance in the distribution of the self-reported responses provided by the participants. Most participants did not use all five points of the five-point Likert scale in their responses, and the distributions tend to skew toward one side or the other, depending on the question. Thus, we binarized the survey data into a two-point scale to give the machine learning models the best possible chance to make useful predictions. The two points in the binary scale represent negative or positive responses for each of the three human aspects of interests (e.g., not stressed or stressed).

@@

\vspace{-8mm}
Even after binarizing the responses as described in the previous section, we found the distribution of responses was still quite imbalanced for many of our participants. This can be seen in the distribution columns in Table \ref{tab:accuracy}. To combat this, we applied random oversampling to our training sets, which artificially rebalances the dataset by creating randomly replicated data in the minority class. This has been a commonly used technique in previous studies on unbalanced datasets \cite{chawla2004,yap2014}.

\begin{table}
  \centering
      \includegraphics[width=0.5\textwidth]{distributiontable.pdf}
  \caption{Distribution of the responses of each participant. Each bar in the histograms represent one of the 5 values on the 5-point Likert scale, the far left side of the histograms are 1/Not at all, and the far right sides are 5/Extremely}
   \label{responseDistribution}
   \vspace*{-2mm}
\end{table}

@@

\subsection{Selecting a Classifier}

\vspace{-4mm}
There are many different algorithms that can be used to build a classifier. To select an
algorithm, we compared multiple classifiers using the popular machine learning library scikit-learn~\cite{pedregosa11} and performed a grid search analysis to determine the optimal hyperparameters for each classifier. Our analysis showed that random forest outperforms all other classifiers, including Na\"ive Bayes, decision trees, support vector machine, and a multilayer perceptron neural network. The optimal values for random forest and the three output measures are listed in Table \ref{tab:hyperparams}. For the remainder of this paper, we refer to random forest classifiers trained with these hyperparameters.%\\[-0.1cm]
% for stress they were (minimum samples for split = 4, # of estimators = 100, max features = 0.5, k=300), for focus they were (minimum samples for split = 4, # of estimators = 50, max features = 0.5, k=200), and for awakeness they were (minimum samples for split = 4, # of estimators = 100, max features = 0.25, k=800)

@@

\vspace{-4mm}
%\noindent\textit{Results of Individual Classifiers}\\
Since peoples' experience of stress, focus, and awakeness (as well as
their physiological manifestations) can vary substantially
(e.g.,~\cite{Hernandez11}), we first trained individual classifiers
for each participant (as opposed to a general one for all
participants). The results of our analysis are reported in
Table~\ref{tab:accuracy}. For our analysis, we report values of
accuracy, one of the most commonly used metric to compare performance,
as well as precision and recall of the classes of interest:
'stressed', 'not focused', and 'not awake'. Since the imbalance in the
data can lead to high accuracy values if a classifier always just
predicts the most likely/frequent class while ignoring the class of
higher importance and interest, precision and recall of the class of
interest are also important to
consider~\cite{yap2014,bhattacharyya_data_2011,Hernandez11}.
For some users (i.e., S1. as seen in Table \ref{responseDistribution}), the imbalance in their data was so extreme that even after adjusting by oversampling it was impossible to create a reasonable classifier. These scenarios are difficult to predict as any classifier will not have enough variance in its training data for the 'stressed' situation to adequately distinguish it from the non-stressed case.

@@

\vspace{-4mm}
%\noindent\textit{Feature Selection and Importance}\\
There are a large variety of features that can be (and have been) calculated in previous research for each of the basic measurements listed in Table~\ref{signals}, such as the mean, standard deviation, maximum, and interquartile range. In addition, each of these metrics can be combined with the various time windows captured of a basic measurement, resulting in a large feature space. To reduce the feature space, we experimented with multiple feature selection methods, including selecting the top k highest correlated features by various metrics such as mutual information, Pearson's correlation coefficient, ANOVA's F-value, as well as wrapper methods such as recursive feature elimination, optimizing mean decrease accuracy by iteratively permuting features, and only selecting features that exceed a certain Gini importance threshold. We found that all methods produced similar results with respect to accuracy, precision, and recall for the individual models. Ultimately, we chose to use the top k features with the highest ANOVA F-value, as it is relatively simple and efficient to calculate. The values of k used were selected by grid search analysis, and are shown in Table \ref{tab:hyperparams}.

@@

\vspace{-4mm}
%\noindent\textit{Individual vs.\ General Model}\\
Individual models are trained specifically for each individual and thus require a data collection period before they are capable of making accurate predictions. On the other hand, the idea of general models is to be able to train them on already collected data and then  to be able to apply them even to new and unseen individuals, thus overcoming the cold-start problem. Given the large individual differences in biometrics, training a general model to achieve an adequate accuracy for new individuals is not necessarily possible.

### 2020-03-26 HICSSSubmission/bp_studyMethod.tex
reducing text

We conducted an eight-week field study with 14 participants using experience sampling and biometric sensors to investigate how knowledge workers experience stress over time and the feasibility of predicting stress, focus, and awakeness based on biometric signals.

@@

We recruited a group of 14 professionals via personal contacts from a large power and automation company. The group is diverse in terms of age, work experience, gender, and work responsibilities. All participants work primarily in an office environment, though half of the participants spend at least 10\% (and up to 50\%) of their time in a laboratory environment. Office workers are a population that generalizes to a variety of contexts, and including part-time laboratory workers guarantees that our participants have varying work patterns that include different levels of computer usage, as well as different levels of activity in both individual and collaborative tasks.

@@

We performed a study over the course of eight weeks. This study includes the collection of the participant's biometric data, computer interaction data, and self-reporting surveys. We informed participants about the study purpose and procedure, handed out biometric sensors, and introduced and explained the self-reporting to the participants. For 10 of the participants, we further installed a computer activity tracker on their company-issued laptops (note: four participants from the original sample declined for privacy reasons).

After the initial setup, participants were asked to fill out 3 surveys each day for the following eight weeks (see Section~\ref{sec:Surveys}). At the end of the eight weeks, we collected the biometric sensors and performed a short follow-up interview on the study and the participants' experiences.

Keeping participants engaged in an eight-week study can be
difficult. In the second month of the study, largely to incent participants to
continue, we offered participants two one-hour Tai Chi classes per
week (each in the middle of the day) and asked them to attend one
class a week. The choice of Tai Chi provided a link to mindfulness,
a topic of growing interest in communities. Consultations we held
with researchers in psychiatry and psychology suggested Tai Chi as
a good link to mindfulness and an intervention that might alleviate
stress. By recording attendence of participants at the Tai Chi
sessions, we are able to analyze the impact of the sessions
on the results. As a brief summary, the impact of the
sessions was minimal; a a result, we present an analysis
of the effects in the Discussion section of the paper.

@@

In this section we describe the two datasets that we collected from each study participant.

@@

Figure~\ref{everion} illustrates Biovotion's Everion, which we used to track the biometric signals of the study participants. The Everion is worn on the upper arm and provides continuous monitoring of certain biometric measurements.\footnote{https://biovotion.zendesk.com/hc/en-us/articles/213613165} Previous studies~\cite{zuger18,sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent} have used similar devices~\cite{Okada11,polar,fitbitCharge} to capture psycho-physiological and biometric measurements for shorter periods of time or capturing a smaller number of measurements.

\begin{figure}
  \centering
      \includegraphics[width=0.4\textwidth]{Everion.jpg}
  \caption{We used Biovotion's Everion to collect biometric measurements from the participants}
   \label{everion}
\end{figure}

Table~\ref{signals} lists the biometrics measurements that we collected using the Everion. Each measurement is collected once per second, and each recorded observation has an associated timestamp and quality rating. Data collected by the Everion are uploaded to a server, from which we downloaded the data for use in our study.

\begin{table}[h!]
\begin{center}
\small\addtolength{\tabcolsep}{-1pt}
\begin{tabular}{l l}
\hline
Biometric Measurement & Units of Measure \\ 
\hline
\textbf{Physical Activity} & \cite{fox1999,aldana1996}\\
\hspace{3mm}Intensity of motion & (No unit)  \\
\hspace{3mm}Energy Expenditure & Calories per second (cal/s)  \\
\hspace{3mm}Step counter & Steps  \\
\hline
\textbf{Heart} & \cite{haapalainen2010psycho,healey2005detecting,mulder1992measurement,haag2004emotion} \\
\hspace{3mm}Heart rate & Beats per Minute (bpm) \\
\hspace{3mm}Blood pulse wave & (No Unit)  \\
\hspace{3mm}Heart rate variability (RMSSD) & Milliseconds (ms) \\
\hspace{3mm}Blood oxygenation  & Percent (\%)  \\
\hspace{3mm}Blood perfusion & (No unit)  \\
\hline
\textbf{Skin} & \cite{healey2005detecting,haag2004emotion} \\
\hspace{3mm}Galvanic skin response &  kOhm  \\
\hspace{3mm}Skin temperature &  Degrees Celsius ($^{\circ}$C)  \\
\hline
\textbf{Respiration} & \cite{mulder1992measurement,healey2005detecting,haag2004emotion,masaoka1997}\\
\hspace{3mm}Respiratory rate & Breaths per Minute (bpm)  \\
\hline
%\textbf{Environmental}\\
%\hspace{3mm}Barometric pressure & Milibar (mbar)  \\
%\hline
\end{tabular}
\caption{Biometric measurements captured by the Everion, organized by category and with references to previous works using similar data}
\label{signals}
\end{center}
%\vspace*{-4mm}
\end{table}

@@

To gain a better understanding of our participants day-to-day work activities, we used an open source computer interaction monitor~\cite{Meyer17}. The monitor ran in the background on participants' computers and tracked the active windows, as well as the keyboard and mouse activity. Four of our participants opted out of this part of the study for privacy reasons (S6, S10, S12, and S13). %Therefore, we only included the computer interaction data in our analysis for our more coarse-grained explanatory model but not in our analysis on the predictive model for stress in the moment.

@@

Following guidelines from previous studies~\cite{Lalle16,Panwar18,Luo18} and 
following the preferrences of extensive user piloting, we sent via 
text message a survey request to each participant two times per 
work day. Pilot participants preferred the usage of text message, in part, due 
to them being accessible and noticeable anywhere in the office.

We sent the 
first request at a random time between 9am and 11am 
and sent the second request at a random time between 1pm and 3pm. We 
randomized the request times to avoid either establishing or observing a 
standard behavioral pattern. That is, we did not want the participants to 
plan for the arrival of the survey request at a set time, and we did not 
want the survey request to overlap with a set daily behavior (e.g., coffee 
break every day at 2:30pm). Similarly, we avoided using tools which allow 
for too much freedom in response time~\cite{Adams18} since this would 
discurage participation in stressed timeframes and would bias the 
corpus. 
%This sentence was added in response to Reviewer 1 comment:
%There is already a lot of work on designing a new self-log, a self-report 
%tool which can blend into the workstation [1] or offers users more freedom 
%in time to self-report [2], or facilitate the self-log by using wearables.
The same survey was sent each time:
\begin{enumerate}
\item How awake are you right now?
\item How stressed do you feel right now?
\item How focused on work are you right now? 
\end{enumerate}
We used the phrase ``right now'' to capture each aspect in the moment (so as to permit later prediction of each aspect based on biometric data). The wordings of the questions are based on a previous survey of individuals in an organizational context~\cite{Gloor_etal:2010}. The use of awakeness (rather than sleepiness) in Question 1 is inspired by previous work~\cite{Wilhelm_Schoebi:2007} and to some extent also captures the ``arousal'' aspect of the affective space~\cite{Russell:1980}.

%\todo{we are not using the last survey for anything in this paper, so I'd leave it out}
One last survey was sent at the end of the day, at 4:25pm which asked the 
four different questions
detailed below:
\begin{enumerate}
\item How awake have you been today?
\item How stressed did you feel today?
\item How productive have you been today?
\item How do you feel about your work day?
\end{enumerate}

>>>>

\vspace{-4mm}
We conducted an eight-week field study with 14 participants using experience sampling and biometric sensors to investigate how knowledge workers experience stress over time and the feasibility of predicting stress, focus, and awakeness based on biometric signals.

@@

\vspace{-4mm}
We recruited a group of 14 professionals via personal contacts from a large power and automation company. The group is diverse in terms of age, work experience, gender, and work responsibilities. All participants work primarily in an office environment, though half of the participants spend at least 10\% (and up to 50\%) of their time in a laboratory environment. Office workers are a population that generalizes to a variety of contexts. Our participants have varying work patterns including varying levels of computer usage, and activity in both individual and collaborative tasks.

@@

\vspace{-4mm}
We performed a study over the course of eight weeks. This study includes the collection of the participant's biometric data, computer interaction data, and self-reporting surveys. We informed participants about the study purpose and procedure, handed out biometric sensors, and introduced and explained the self-reporting to the participants. For 10 of the participants, we further installed a computer activity tracker on their company-issued laptops (note: four participants from the original sample declined for privacy reasons).

After the initial setup, participants were asked to fill out three surveys each day for the following eight weeks (see Section~\ref{sec:Surveys}). At the end of the eight weeks, we collected the biometric sensors and performed a short follow-up interview on the participants' experiences.

Keeping participants engaged in an eight-week study can be
difficult. In the second month of the study, largely to incent participants to
continue, we offered participants two one-hour Tai Chi classes per
week (each in the middle of the day) and asked them to attend one
class a week. The choice of Tai Chi provided a link to mindfulness,
a topic of growing interest in communities. Consultations we held
with researchers in psychiatry and psychology suggested Tai Chi as
a good link to mindfulness and an intervention that might alleviate
stress. By recording attendance of participants at the Tai Chi
sessions, we are able to analyze the impact of the sessions
on the results.

@@

\vspace{-4mm}
In this section we describe the two datasets that we collected from each study participant.

@@

\vspace{-8mm}
Figure~\ref{everion} illustrates Biovotion's Everion, which we used to track the biometric signals of the study participants. The Everion is worn on the upper arm and provides continuous monitoring of certain biometric measurements.\footnote{https://biovotion.zendesk.com/hc/en-us/articles/213613165} Previous studies~\cite{zuger18,sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent} have used similar devices~\cite{Okada11,polar,fitbitCharge} to capture psycho-physiological and biometric measurements for shorter periods of time or capturing a smaller number of measurements.

\begin{figure}
  \centering
      \includegraphics[width=0.4\textwidth]{Everion.jpg}
  \caption{Biovotion's Everion}
   \label{everion}
\end{figure}

Table~\ref{signals} lists the biometric measurements that we collected using the Everion. Each measurement is collected once per second, and each recorded observation has an associated timestamp and quality rating. Data collected by the Everion is uploaded to a server, from which we downloaded the data for our study.

\begin{table}[h!]
\begin{center}
\small\addtolength{\tabcolsep}{-1pt}
\begin{tabular}{l l}
\hline
Biometric Measurement & Units of Measure \\ 
\hline
\textbf{Physical Activity} & \cite{fox1999,aldana1996}\\
\hspace{3mm}Intensity of motion & (No unit)  \\
\hspace{3mm}Energy Expenditure & Calories per second (cal/s)  \\
\hspace{3mm}Step counter & Steps  \\
\hline
\textbf{Heart} & \cite{haapalainen2010psycho,healey2005detecting,mulder1992measurement,haag2004emotion} \\
\hspace{3mm}Heart rate & Beats per Minute (bpm) \\
\hspace{3mm}Blood pulse wave & (No Unit)  \\
\hspace{3mm}Heart rate variability (RMSSD) & Milliseconds (ms) \\
\hspace{3mm}Blood oxygenation  & Percent (\%)  \\
\hspace{3mm}Blood perfusion & (No unit)  \\
\hline
\textbf{Skin} & \cite{healey2005detecting,haag2004emotion} \\
\hspace{3mm}Galvanic skin response &  kOhm  \\
\hspace{3mm}Skin temperature &  Degrees Celsius ($^{\circ}$C)  \\
\hline
\textbf{Respiration} & \cite{mulder1992measurement,healey2005detecting,haag2004emotion,masaoka1997}\\
\hspace{3mm}Respiratory rate & Breaths per Minute (bpm)  \\
\hline
%\textbf{Environmental}\\
%\hspace{3mm}Barometric pressure & Milibar (mbar)  \\
%\hline
\end{tabular}
\caption{Biometric measurements captured by the Everion}
\label{signals}
\end{center}
%\vspace*{-4mm}
\end{table}

@@

\vspace{-8mm}
To gain a better understanding of our participants day-to-day work activities, we used an open source computer interaction monitor~\cite{Meyer17}. The monitor ran in the background on participants' computers and tracked the active windows, as well as the keyboard and mouse activity. Four of our participants opted out of this part of the study for privacy reasons (S6, S10, S12, and S13). %Therefore, we only included the computer interaction data in our analysis for our more coarse-grained explanatory model but not in our analysis on the predictive model for stress in the moment.

@@

\vspace{-8mm}
Following guidelines from previous studies~\cite{Lalle16,Panwar18,Luo18} and 
following the preferrences of extensive user piloting, we sent via 
text message a survey request to each participant two times per 
work day. Pilot participants preferred the usage of text message, in part, due 
to them being accessible and noticeable anywhere in the office.

We sent the 
first request at a random time between 9am and 11am 
and sent the second request at a random time between 1pm and 3pm. We 
randomized the request times to avoid either establishing or observing a 
standard behavioral pattern. That is, we did not want the participants to 
plan for the arrival of the survey request at a set time, and we did not 
want the survey request to overlap with a set daily behavior (e.g., coffee 
break every day at 2:30pm). Similarly, we avoided using tools which allow 
for too much freedom in response time~\cite{Adams18} since this would 
discurage participation in stressed timeframes and would bias the 
corpus. 
%This sentence was added in response to Reviewer 1 comment:
%There is already a lot of work on designing a new self-log, a self-report 
%tool which can blend into the workstation [1] or offers users more freedom 
%in time to self-report [2], or facilitate the self-log by using wearables.
The same survey was sent each time:
\begin{enumerate}[topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
\item How awake are you right now?
\item How stressed do you feel right now?
\item How focused on work are you right now? 
\end{enumerate}
We used the phrase ``right now'' to capture each aspect in the moment (so as to permit later prediction of each aspect based on biometric data). The wordings of the questions are based on a previous survey of individuals in an organizational context~\cite{Gloor_etal:2010}. The use of awakeness (rather than sleepiness) in Question 1 is inspired by previous work~\cite{Wilhelm_Schoebi:2007} and to some extent also captures the ``arousal'' aspect of the affective space~\cite{Russell:1980}.

%\todo{we are not using the last survey for anything in this paper, so I'd leave it out}
One last survey was sent at the end of the day, at 4:25pm which asked the 
four different questions
detailed below:
\begin{enumerate}[topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
\item How awake have you been today?
\item How stressed did you feel today?
\item How productive have you been today?
\item How do you feel about your work day?
\end{enumerate}
